Child survival for mothers: mortality change and related measures

### 2020-04-30 ALAP2020/ALAP.Rmd
sugerencias antes de charla el viernes

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[This runs contrary to the findings of @KeyfitzMath, who found that levels of fertility did not affect the probability of having living ancestors.] and the mean age at childbearing $\kappa$ acts through the factor $l_{a-\kappa}$.
The dispersion of of fertility $\sigma^2$ is negatively proportional to child survival. 
All else constant, a higher dispersion of fertility $\sigma^2$ is associated with a decrease in $CS_a$ during reproductive age.
The negative relationship stems from the fact that the second derivative of $l_{a-\kappa}$ is negative in the 15-49 age range, given a concave downwards survival function.

@@

### Burden and timing

@@

```{r plot_ITL_MAL, fig.cap = "Intensity Tome Lost of  womans aged 30, 40 and 50. Years 1950 (darker) to 2015 (lighter). Latinamerican countries in period 1950-2015"}
ITL <- ggplot(CS_outs)+  
                    geom_smooth(method="lm", aes(x=Fa, y=ITL, color=factor(Age)), 
                                se = F, formula= (y ~ exp(x)))+
                    geom_point(aes(x=Fa, y=ITL, color=factor(Age)), alpha = .3) +
                    theme_classic()+        
                    theme(legend.position = "right") +
                    labs(y = "%")

@@

We now consider the consequences of an absolute change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk. Considering that $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012), and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:

@@

```{r CS_abs_app, fig.cap= "Absolute change and its approximation for age CS in women aged 50. Latinamerican countries in period 1950-2015"}
# application of absolute changes
age = 50
CS_abs_change_app = data.frame(delta = NA,Country = NA, Period = NA, change = 0, change_app = 0)
for(delta in seq(0,.01,.001)){
  for (country in unique(LA$name)){
    for (period in unique(LA$Period)){
      # country = "Argentina"; period = "2010-2015"; delta = .01
      change_app = -(age - CS_outs %>% filter(name == country, 
                                              Period == period, 
                                              Age == age) %>% select(k)) * delta
      country_i = LA %>% filter(name == country, Period == period) 
      CS = CS_outs %>% filter(name == country, 
                              Period == period, 
                              Age == age) %>% select(CS)
      CS_changed = sum(unlist(sapply(1:age, function(i) 
        country_i$asfr[i]*country_i$lx[age-(i-1)]/100000*exp(-delta*(age-(i-1))))), 
        na.rm = T)
      change = (CS_changed-CS)/CS
      CS_abs_change_app = rbind(CS_abs_change_app, 
                                data.frame(delta = delta,
                                           Country = country, Period = period, 
                                           change = as.numeric(change), 
                                           change_app = as.numeric(change_app)))
    }
  } 
}
CS_abs_change_app %>% slice(-1) %>% 
  ggplot() + 
  geom_point(aes(x=change, y=change_app, color=delta)) +
  geom_abline(slope=1, intercept = 0) +
  theme_classic() +
  # theme(legend.position = "none")+
  xlim(-.3, 0) + ylim(-.3, 0)
```

That an equal change in mortality could happens at all ages is unlikelihood. As a [corollary](#Corollary) we inspect the case that only infant mortality is affected.

The relationship is less clear for a proportional change given the interaction of birth and mortality rates (@KeyfitzMath). Considering that $\mu_{x,\delta}=\mu_{x}(1+\delta)$ and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$:

$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx$$

@@

This factor has a S-shape because $l_x$ curvature fertility accumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no additional children and higher risk for one age additional of survive).

@@

In a model with heterogeneity, determined at born with a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *K* for fertility heterogeneity and *Z* for mortality frailty.

@@

Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline accumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$). Frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}Z}=l_x^Z$, with baseline hazard $\mu$. Replacing conveniently the last two expressions and considering the joint distribution $f_{kz}=f_{z/k} \,f_k$, the mean $\overline{CS}_a$ would be:

>>>>

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[Note that this runs contrary to the findings of @KeyfitzMath, who found that levels of fertility did not affect the probability of having living ancestors.] and the mean age at childbearing $\kappa$ acts through the factor $l_{a-\kappa}$.
The dispersion of of fertility $\sigma^2$ is negatively proportional to child survival. 
All else constant, a higher dispersion of fertility $\sigma^2$ is associated with a decrease in $CS_a$ during reproductive age.
The negative relationship stems from the fact that the second derivative of $l_{a-\kappa}$ is negative in the 15-49 age range, given a concave downwards survival function.

@@

### Burden and timing of maternal bereavement

@@

```{r plot_ITL_MAL, fig.cap = "Intensity Tome Lost of  womans aged 30, 40 and 50. Years 1950 (darker) to 2015 (lighter). Latinamerican countries in period 1950-2015"}
ITL <- 
  ggplot(CS_outs) +
  geom_smooth(
    method="lm"
    , aes(x=Fa, y=ITL, color=factor(Age))
    , se = F
    , formula= (y ~ exp(x))
    ) +
  geom_point(aes(x=Fa, y=ITL, color=factor(Age)), alpha = .3) +
  theme_classic() +
  theme(legend.position = "right") +
  labs(y = "%")

@@

We now consider the consequences of an absolute change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk. 
As a [corollary](#Corollary) we consider the case where mortality only changes for infants, considering that it is unlikelily for a potential change in mortality to affect all ages equally in real populations.
For now, let us consider $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012) and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:

@@

```{r CS_abs_app, fig.cap= "Effect of an absolute absolute change in mortality on child survival by maternal age of women in Latin America  for the 1950-2015 period. Goodness of fit of approximation compared to direct estimation"}

# NEW CODE ----

abs30 <- absolute_change_fit(30)
abs40 <- absolute_change_fit(40)
abs50 <- absolute_change_fit(50)
abs70 <- absolute_change_fit(70)

bind_rows(abs30, abs40, abs50, abs70) %>% 
  mutate(age = paste("Maternal age", age)) %>% 
  ggplot() + 
  geom_point(aes(x=change, y=change_app, color=delta)) +
  geom_abline(slope=1, intercept = 0) +
  scale_x_continuous("Actual change in CS", limits = c(-.3, 0)) +
  scale_y_continuous("Approximated change in CS", limits = c(-.3, 0)) +
  scale_color_continuous("Absolute change\nin delta") +
  facet_wrap(~age) +
  theme_classic()

# ORIGINAL CODE ----
# application of absolute changes

# age <- 50
# 
# CS_abs_change_app <- data.frame(delta = NA,Country = NA, Period = NA, change = 0, change_app = 0)
# 
# for(delta in seq(0,.01,.001)){
# # for(delta in seq(0,.02,.005)){
#   for (country in unique(LA$name)){
#     for (period in unique(LA$Period)){
#       # country = "Argentina"; period = "2010-2015"; delta = .01
#       change_app = -(age - CS_outs %>% filter(name == country, 
#                                               Period == period, 
#                                               Age == age) %>% 
#                        select(k)) * delta
#       
#       country_i = LA %>% filter(name == country, Period == period) 
#       CS = CS_outs %>% filter(name == country, 
#                               Period == period, 
#                               Age == age) %>% select(CS)
#       CS_changed = sum(unlist(sapply(1:age, function(i) 
#         country_i$asfr[i]*country_i$lx[age-(i-1)]/100000*exp(-delta*(age-(i-1))))), 
#         na.rm = T)
#       change = (CS_changed-CS)/CS
#       CS_abs_change_app = rbind(CS_abs_change_app, 
#                                 data.frame(delta = delta,
#                                            Country = country, Period = period, 
#                                            change = as.numeric(change), 
#                                            change_app = as.numeric(change_app)))
#     }
#   } 
# }
# 
# CS_abs_change_app %>% 
#   slice(-1) %>% 
#   ggplot() + 
#   geom_point(aes(x=change, y=change_app, color=delta)) +
#   geom_abline(slope=1, intercept = 0) +
#   scale_x_continuous("Actual change in CS", limits = c(-.3, 0)) +
#   scale_y_continuous("Approximated change in CS", limits = c(-.3, 0)) +
#   scale_color_continuous("Absolute change\nin delta") +
#   theme_classic()

```

We now consider the consequences of a proportional change in mortality on child survival.
Given a proportional change in mortality $\mu_{x,\delta}=\mu_{x}(1+\delta)$ so that $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$, it follows that:

$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx.$$

@@

Grasping this relationship intuitively is more difficult to  given the interaction of birth and mortality rates (@KeyfitzMath). 
This factor has a S-shape because $l_x$ curvature fertility accumulation, that gives more weight to first ages, and which is grater than 1 only for ages *a* with complete reproductive life and high infant mortality (with no additional children and higher risk for one age additional of survive).

<!-- Do we get any inuition about how proportional changes affects CS? -->

@@

In a model with heterogeneity, determined at birth with a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *K* for fertility heterogeneity and *Z* for mortality frailty.

@@

Following @Coresh1988 we can express the fertility component in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline accumulated fertility, $K$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$). Frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}Z}=l_x^Z$, with baseline hazard $\mu$. Replacing conveniently the last two expressions and considering the joint distribution $f_{kz}=f_{z/k} \,f_k$, the mean $\overline{CS}_a$ would be:
